\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-03 chapters/bci.tex
Resturcture chap 1 and 2

\todo{comments liesel paradigms})
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}

@@

In general, devices processing direct inputs from the central nervous system
are useful for rehabilitation and medical diagnosis or treatment.
Additionally, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality.\todo{cite}
Ultimatly, they have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
It is this \ac{bci} controlled communication is where our focus lies,
as it offers a transformative means for individuals with severe motor
impairments to regain functional independence to write and speak through
assistive technology.
This allows them to talk to their loved ones and caregivers, exercise hobbies
and makes many aspects of daily life more accessible.

\section{A direct interface to the brain}
\todohsm{Start with state of the art before theses.}
\todohsm{objectives at the end of chapter 1}
\todohsm{Introduction part: chapter 1 rationale, chapter 2: literature/sota}
\todohsm{make less verbose}

@@

\fullpagefigct{%
  \includegraphics[width=\textwidth]{figures/bci/bci_loop.pdf}
}{%
  The \ac{bci} loop.
}{%
  The \ac{bci} loop.
  The user interacts with the \ac{bci} via a specific paradigm, in this case involving
  visual stimulation.
  Electrical neural signals are recorded, and neurophysiological features
  related to the paradigm are extracted. Using machine learning, a decision can
  then be made based on these features, which can be presented back to the user.
  In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{%
  fig:bci/loop
}

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what is going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to establish the
activity of a specific brain region
However, this signal reacts too slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential}
is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical activity are be faster and more
informative.

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
A \ac{bci} operates according to a specified paradigm, which defines the method by which interaction is conducted.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the inteded letter A or B
in Figure~\ref{fig:bci/loop}.
The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific
brain signatures related to our paradigm need to be identified within the recorded
activity.
This is commonly referred to as \emph{feature extraction}.
In other words, the signal needs to be interpreted \emph{in function of} the
task we want our \ac{bci} to perform.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own brain activity.
On the other hand, the activity of interest often
originates from a specific region or network within the brain.
This activity might not easily be discerned from all other brain and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing} (see Section~\ref{sec:bci/preprocessing}), but the problem of retrieving or \emph{decoding} only the
relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder.
Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.
For a communication system, this means coupling the decoded information to
e.g.\ a virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter
the user's cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like tactile
feedback for movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural
pathways, can have a positive impact on \ac{bci} performance and gives rise to
rehabilitation applications.
\todo{is every section referred to?}

\todo{wrap-up}
\todo{make shorter}

\section{Recording technologies}
\label{sec:bci-recording}
\todo{why is spatial resolution a good thing}
\todo{explain with each method what it actually is (e.g.\ electrodes placed
at...)}
\todo{make sure every part ends with a short summary motivation for why we chose our
approach}

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
The technologies above all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality of electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials and
postsynaptic potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/bci/recording_modalities.pdf}
}{%
  Various electrophysiology recording technologies.
}{%
	Various electrophysiology recording technologies and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{%
  fig:bci/recording
}

Specifically, invasive methods come with an increased \emph{spatial resolution},
meaning they can extract a signal from a specific brain region or even a set of
neurons of interest.
Generally, penetrating microelectrodes are considered to have the highest
spatial resolution.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular
potential of a small group of neurons, or even intracellular single neuron
action potential spikes.
They come in single-electrode form, or, more popularly, in microelectrode
arrays.
Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
NeuraLink implant~\cite{Musk2019} and IMEC's Neuropixels
2.0~\cite{Steinmetz2021}.

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
have been realized using
invasive mictroelectrode arrays~\cite{Willett2021} and $\mu$\Ac{ecog}\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution~\cite{Steinmetz2021}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
through the bloodstream via a catheter, removing the need for open brain
surgery.
Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
technology is preferred over implanted electrodes, unless the added value of invasive BCI is
sufficiently large~\cite{Kageyama2020}.\todo{due to what reasons}
This provides justification for focusing part of the current research effort on
non-invasive BCI.

@@

Despite its noisy nature, \ac{eeg} is the recording methodology of choice for
our \ac{bci} because of its wide acceptance by patients.
To counteract the noise present in our recording, we must on the one hand evoke
strong, informative brain signals using a suited \ac{bci} paradigm, and on the
other hand pick a suited decoder which can isolate the signal of interest and
filter out the noise.

@@

\section{Preprocessing}
\label{sec:bci/preprocessing}
\todo{preprocessing}
EEG preprocessing is a critical step in brain-computer interface (BCI) systems, as it
significantly improves the quality of the recorded data for subsequent analysis and
classification. Common preprocessing techniques in BCI systems include
rereferencing, band-pass and notch filtering, and \ac{ica}.

@@

\subsection{The \acf{erp}}

@@

\subsection{Preprocessing}
In addition to (ref) epoching
baseline correction, but not in decoding (cite)

@@

>>>>

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\the\textwidth

@@

In general, devices processing direct inputs from the central nervous system
are useful for rehabilitation and medical diagnosis or treatment.
Additionally, they bring a new paradigm to human-computer interaction,
especially when paired with virtual or augmented reality~\cite{SiMohammed2017}.
Ultimatly, they have always been especially promising as assistive technology for
paralyzed patients limited in their communication ability.
It is this \ac{bci} controlled communication is where our focus lies,
as it offers a transformative means for individuals with severe motor
impairments to regain functional independence to write and speak through
assistive technology.
This allows them to talk to their loved ones and caregivers, exercise hobbies
and makes many aspects of daily life more accessible.

\section{A direct interface to the brain}
\todohsm{make less verbose}

@@

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/bci_loop.pdf}
}{%
  The \ac{bci} loop.
}{%
  The \ac{bci} loop.
  The user interacts with the \ac{bci} via a specific paradigm, in this case involving
  visual stimulation.
  Electrical neural signals are recorded, and neurophysiological features
  related to the paradigm are extracted. Using machine learning, a decision can
  then be made based on these features, which can be presented back to the user.
  In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{%
  fig:bci/loop
}

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what is going on in the brain.
Multiple signals can be recorded from the brain, including bloodflow,
magnetical electrical activity.
For several reasons laid out in~\ref{sec:bci-recording}, the electrical modality
is especially interesting.

The recorded brain activity forms only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
Instead, it would help greatly if we knew which specific activity to look for.
A \ac{bci} operates according to a specified \emph{paradigm}, which defines the
method by which interaction is conducted.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
to couple specific activity in the brain to specific information or actions we
want to convey via the \ac{bci}.
The brain singal now \emph{modulates} this information.
In turn, the detection of this signal can be coupled to an action, like
typing the inteded letter A or B in Figure~\ref{fig:bci/loop}.
The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

The previous points to another component of a successful \ac{bci}: the specific
brain signatures related to our paradigm need to be identified within the recorded
activity.
The electrical activity of the brain is relatively weak compared to that of the
environment around us, containing electronic devices and other sources of
interference.
Furthermore, the brain is continuously processing information and carrying out
`background tasks', all of which generate their own brain activity.
On the other hand, the activity of interest often
originates from a specific region or network within the brain.
This activity might not easily be discerned from all other brain and
environmental activity.
Compare this to trying to listen to a single speaker at a noisy party where
everyone is shouting over each other.
Some obvious sources of interference can be easily filtered out by
\emph{preprocessing} (see Section~\ref{sec:bci/preprocessing}) and by choosing
the correct representation of the signal (i.e. \emph{feature extraction}, but the problem
of retrieving or \emph{decoding} only the relevant signal is generally hard.
Nevertheless, we can attempt to solve it through supervised machine learning.
Useful information that is encoded in the brain signal by the paradigm,
can be retrieved from it by the decoder and interpreted in function of the
desired \ac{bci} output.
Section~\ref{sec:bci/decoding} takes a closer look at some common decoding
techniques.

Finally, the loop can be closed by coupling this to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.
For a communication system, this means coupling the decoded information to
e.g.\ a virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter
the user's cognitive state or brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as a hearing prosthesis.
More indirectly, modulation can be achieved by sensory stimulation.
This involves specific, paradigm-related sensory input, like tactile
feedback for movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural
pathways, can have a positive impact on \ac{bci} performance and gives rise to
rehabilitation applications.

To summarize, a \ac{bci} can facilitate direct interaction central nervous system.
To do this, information is modulated in the brain signal according to the
\ac{bci} paradigm by stimulating, the user or having them perform a task.
Their brain signal is subsequently recorded and preprocessed to clean it.
Relevant features that contain aspects of the modulated signal are extracted,
and the modulate information is retrieved by the decoder, a machine learning
model.
This information can then acted on and the loop is closed if this action
affects the user.
Let us take a look at each of these elements separately in the next section.

\section{Recording technologies}
\label{sec:bci-recording}

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.

The technologies above all measure bloodflow (the hemodynamic signal), which only indirectly
reflects brain activity.
This signal reacts too slowly to reflect the real-time activity that is
of interest in \ac{bci} for communication, and it carries little information
other than the brain region from where it is originating.

A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential}
is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
The related discipline is \emph{electrophysiology}.

Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality of electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials and
postsynaptic potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

@@

\fullpagefig[topleft]{%
  \includegraphics[width=\fullpagefigwidth]{figures/bci/recording_modalities.pdf}
}{%
  Various electrophysiology recording technologies.
}{%
	Various electrophysiology recording technologies and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{%
  fig:bci/recording
}

Specifically, invasive methods are less noisy and come with an increased \emph{spatial resolution},
meaning they can extract a signal from a specific brain region or even a set of
neurons of interest.
This makes it easier to retrieve a signal with known origin from the brain,
which might be leveraged by the paradigm to modulate information.
Generally, microelectrodes penetrating the cortex are considered to have the highest
spatial resolution.
These are needles with a diameter smaller than 10µm and one or multiple
electrodes that extend several millimmeters into the cortex.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular
potential of a small group of neurons, or even intracellular single neuron
action potential spikes.
They come in single-electrode form, or, more popularly, in microelectrode
arrays.
Well-known examples are the Utah array~\cite{Maynard1997} and the more recent
NeuraLink implant~\cite{Musk2019} and IMEC's Neuropixels
2.0~\cite{Steinmetz2021}.

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Depth electrodes are longer rods with multiple electrodes than can extend
deeper into the brain.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
have been realized using
invasive mictroelectrode arrays~\cite{Willett2021} and $\mu$\Ac{ecog}\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution~\cite{Steinmetz2021}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
through the bloodstream via a catheter, removing the need for open brain
surgery.
Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
technology is preferred over implanted electrodes, unless the added value of invasive BCI is
sufficiently large~\cite{Kageyama2020}.
This provides justification for focusing part of the current research effort on
non-invasive BCI.

@@

Despite its noisy nature, \ac{eeg} is the recording methodology of choice for
our \ac{bci} because of its wide acceptance by patients.
To counteract the noise present in our recording, we must on the one hand evoke
strong, informative brain activity using a suited \ac{bci} paradigm, and on the
other hand pick a suited decoder which can isolate the signal of interest and
filter out the noise.

@@

\section{Preprocessing}%
\label{sec:bci/preprocessing}
\todo{preprocessing}
EEG preprocessing is a critical step in brain-computer interface (BCI) systems, as it
significantly improves the quality of the recorded data for subsequent analysis and
classification. Common preprocessing techniques in BCI systems include
rereferencing, band-pass and notch filtering, and \ac{ica}.

@@

\subsection{The \acf{erp}}
\label{sec:bci/oddball/erp}

@@

\subsection{Preprocessing \& feature extraction}
In addition to (ref) epoching
baseline correction, but not in decoding (cite)

@@

\todo{wrap-up}

\section{The gaze-dependence problem}

In a nutshell, we can state that \acp{bci} decode brain activity with the aim to
establish a direct communication pathway bypassing speech and other forms of
muscular activity~\cite{Naci2012,Chaudhary2016}.
They have raised great hopes for individuals devoid of these abilities, for whom
\acp{bci}s can provide a means to communicate or to
control devices.

Furthermore, the \ac{eeg}-based visual event-related potential (\ac{erp})-based
oddball interface is an effective and proven method~\cite{Wolpaw2018,Severens2020}.
Targets are shown in short flashes on a computer screen, that evoke
and \ac{erp} when observed, which can be detected in the \ac{eeg}-signal
detected in the EEG-signal.
The \ac{erp} consists of multiple components, which some are modulated by
perception and the attention of the user.
Decoding this modulation allows for information transfer controlled by the
user's brain activity, as sketched in Figure~\ref{fig:bci/loop}.

However, when carefully examining the properties of the \ac{erp} components,
we notice that while attention plays a role in the modulation of specific components,
some part is also related to visual perception.
How a stimulus is perceived, and, as a consequence, which \ac{erp} components
are modulated, is directly related to whether the user gazes directly at it
(i.e.\ \emph{fixates} their gaze on the target) or not.
Indeed, research has shown that visual oddball \ac{bci}s cannot operate
efficiently when the user does not direct
their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.

This leads to a problem that is not often addressed:
a visual \acp{bci} is dependent on the gaze of the user, and hence on
muscle control.
This is exactly what \acp{bci} for augmented and assistive technology try to
avoid.
The \ac{bci} target population consists of individuals who suffer from various
degrees of paralysis, or even from \ac{lis}, the complete loss of muscle
control with preserved consciousness.
While \ac{bci}s are most attractive as a solution for individuals with
\ac{sspgi}, studies
paradoxically often report bad performance in patient populations precisely due
to the lack of adequate eye motor control.
Equally often, participants with gaze impairments were excluded from the study.

Different degrees of eye motor impairment can render visual \ac{bci}s
uncomfortable or outright impossible to use.
When operating a visual \ac{bci}, the usual rapid series of forced saccades
followed by fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest
position of their choice while operating the \ac{bci}.
This points to the need for gaze-independent solutions, which we will explore
further in Chapter~\ref{sec:gaze-independence} and the rest of this thesis.

### 2024-10-03 chapters/bttda.tex
Resturcture chap 1 and 2

To apply HODA in a classification setting, the projections
are first learned on a training dataset  with known class labels.
Next, these projections are used to extract latent tensors from the
tensors in the training dataset.
These latent training tensors are then reshaped (\emph{vectorized}) into feature vectors
$\vec{g} =  \vect(\ten{G})$ and used to train a decision classifier with the corresponding class labels.
At the evaluation stage, the projections learned from the training dataset are
used to extract latent tensors from an unseen test dataset with unknown class
labels, which can also be vectorized and passed on to the trained decision
classifier.

@@

The resulting latent tensors can be vectorized and concatenated into
one single feature vector per input tensor
\begin{equation}
	\vec{g}
	=\left[\vect\left(\ten{G}^{(1)}\right)\
		\vect\left(\ten{G}^{(2)}\right)\
		\cdots\
		\vect\left(\ten{G}^{(B)}\right)\right]
\end{equation}
so that they can be classified in a similar manner to HODA.

>>>>

To apply HODA in a classification setting, the projections
are first learned on a training dataset  with known class labels.
Next, these projections are used to extract latent tensors from the
tensors in the training dataset.
These latent training tensors are then reshaped (\emph{vectorized}) into feature vectors
$\mat{g} =  \vect(\ten{G})$ and used to train a decision classifier with the corresponding class labels.
At the evaluation stage, the projections learned from the training dataset are
used to extract latent tensors from an unseen test dataset with unknown class
labels, which can also be vectorized and passed on to the trained decision
classifier.

@@

The resulting latent tensors can be vectorized and concatenated into
one single feature vector per input tensor
\begin{equation}
	\mat{g}
	=\left[\vect\left(\ten{G}^{(1)}\right)\
		\vect\left(\ten{G}^{(2)}\right)\
		\cdots\
		\vect\left(\ten{G}^{(B)}\right)\right]
\end{equation}
so that they can be classified in a similar manner to HODA.
